Comparison of Terrain Following and Cut Cell Grids using a Non-Hydrostatic Model

### 2016-01-12 src/tf-cutcell-comparison/WellerAppendix.tex

The semi-implicit technique involves combining the momentum \ref{eq:momentum},
continuity \ref{eq:cont} and $\theta$ \ref{eq:theta} equations
and the equation of state \ref{eq:state} to form a Helmholtz equation
to be solved implicitly, as described by \citet{weller-shahrokhi2014}.
The semi-implicit solution technique with a Hodge operator can be
defined by considering only a discretised form of the continuity equation:
\begin{equation}
\frac{\phi^{(n+1)}-\rho^{(n)}}{\Delta t}+\frac{1}{2}\left\{ \nabla\cdot(HV)^{(n)}+\nabla\cdot(HV)^{(n+1)}\right\} =0.\label{eq:contD}
\end{equation}
The divergence is discretised using Gauss' divergence theorem so that:
\begin{equation}
\nabla\cdot\left(HV\right)=\frac{1}{\mathcal{V}_{c}}\sum_{f\in c}n_{f}(HV)_{f}\label{eq:GaussDiv}
\end{equation}
where $\mathcal{V}_{c}$ is the volume of cell $c$, $f\in c$ denotes
the faces of cell $c$, and $n_{f}=1$ if $\mathbf{d}_{f}$ points
outwards from the cell and $n_{f}=$-1 otherwise. Equation \ref{eq:GaussDiv}
is now a sum over a sum since $(HV)_{f}$ is one element of a matrix-vector
multiply. In order to simplify the construction of the matrix for
the Helmholtz problem, only the diagonal terms of $HV$ are treated
implicitly. Therefore, $H$ is separated into a diagonal and off-diagonal
matrix:
\begin{equation}
H=H_{d}+H_{off}.
\end{equation}
Equation \ref{eq:contD} can now be approximated by:
\begin{equation}
\frac{\phi^{(n+1)}-\rho^{(n)}}{\Delta t}+\frac{1}{2}\left\{ \nabla\cdot(HV)^{(n)}+\nabla\cdot(H_{d}V)^{(n+1)}+\nabla\cdot(H_{off}V)^{\ell}\right\} =0\label{eq:contDd}
\end{equation}
where superscript $\ell$ denotes lagged values taken from a previous
iteration or from a previous stage of a Runge-Kutta scheme. This was
the approach taken by \citet{weller-shahrokhi2014}. However, the
numerical solution of equation \ref{eq:contDd} turns out to be unstable
when using a large time-step on highly non-orthogonal meshes associated
with terrain following layers over steep orography. Improved stability
and energy conservation can be achieved by splitting $H$ into a diagonal
component which would be correct on an orthogonal grid and a non-orthogonal
correction:
\begin{equation}
H=H_{c}+H_{corr}
\end{equation}
where the diagonal matrix $H_{c}=|\mathbf{S}_{f}|/|\mathbf{d}_{f}|$
and the non-orthogonal correction is $H_{corr}=H-H_{c}$. The orthogonal
part, $H_{c}$, can be treated implicitly in the Helmholtz equation:
\begin{equation}
\frac{\phi^{(n+1)}-\rho^{(n)}}{\Delta t}+\frac{1}{2}\left\{ \nabla\cdot(HV)^{(n)}+\nabla\cdot(H_{c}V)^{(n+1)}+\nabla\cdot(H_{corr}V)^{\ell}\right\} =0.\label{eq:contDc}
\end{equation}
This form is used for the solutions of the Euler equations in this
paper and is stable, with good energy conservation for all of the
tests presented.

>>>>

The semi-implicit technique involves combining the momentum (\ref{eq:momentum}),
continuity (\ref{eq:cont}) and $\theta$ (\ref{eq:theta}) equations
and the equation of state (\ref{eq:state}) to form a Helmholtz equation
to be solved implicitly, as described by \citet{weller-shahrokhi2014}.
The semi-implicit solution technique with a Hodge operator can be
defined by considering only a discretised form of the continuity equation:
\begin{equation}
\frac{\phi^{(n+1)}-\rho^{(n)}}{\Delta t}+\frac{1}{2}\left\{ \nabla\cdot(HV)^{(n)}+\nabla\cdot(HV)^{(n+1)}\right\} =0.\label{eq:contD}
\end{equation}
The divergence is discretised using Gauss' divergence theorem so that:
\begin{equation}
\nabla\cdot\left(HV\right)=\frac{1}{\mathcal{V}_{c}}\sum_{f\in c}n_{f}(HV)_{f}\label{eq:GaussDiv}
\end{equation}
where $\mathcal{V}_{c}$ is the volume of cell $c$, $f\in c$ denotes
the faces of cell $c$, and $n_{f}=1$ if $\mathbf{d}_{f}$ points
outwards from the cell and $n_{f}=$-1 otherwise. Equation (\ref{eq:GaussDiv})
is now a sum over a sum since $(HV)_{f}$ is one element of a matrix-vector
multiply. In order to simplify the construction of the matrix for
the Helmholtz problem, only the diagonal terms of $HV$ are treated
implicitly. Therefore, $H$ is separated into a diagonal and off-diagonal
matrix:
\begin{equation}
H=H_{d}+H_{off}.
\end{equation}
Equation (\ref{eq:contD}) can now be approximated by:
\begin{equation}
\frac{\phi^{(n+1)}-\rho^{(n)}}{\Delta t}+\frac{1}{2}\left\{ \nabla\cdot(HV)^{(n)}+\nabla\cdot(H_{d}V)^{(n+1)}+\nabla\cdot(H_{off}V)^{\ell}\right\} =0\label{eq:contDd}
\end{equation}
where superscript $\ell$ denotes lagged values taken from a previous
iteration or from a previous stage of a Runge-Kutta scheme. This was
the approach taken by \citet{weller-shahrokhi2014}. However, the
numerical solution of equation (\ref{eq:contDd}) turns out to be unstable
when using a large time-step on highly non-orthogonal meshes associated
with terrain following layers over steep orography. Improved stability
and energy conservation can be achieved by splitting $H$ into a diagonal
component which would be correct on an orthogonal grid and a non-orthogonal
correction:
\begin{equation}
H=H_{c}+H_{corr}
\end{equation}
where the diagonal matrix $H_{c}=|\mathbf{S}_{f}|/|\mathbf{d}_{f}|$
and the non-orthogonal correction is $H_{corr}=H-H_{c}$. The orthogonal
part, $H_{c}$, can be treated implicitly in the Helmholtz equation:
\begin{equation}
\frac{\phi^{(n+1)}-\rho^{(n)}}{\Delta t}+\frac{1}{2}\left\{ \nabla\cdot(HV)^{(n)}+\nabla\cdot(H_{c}V)^{(n+1)}+\nabla\cdot(H_{corr}V)^{\ell}\right\} =0.\label{eq:contDc}
\end{equation}
This form is used for the solutions of the Euler equations in this
paper and is stable, with good energy conservation for all of the
tests presented.

### 2016-01-12 src/tf-cutcell-comparison/fig-gravityWaves-theta.tex
discuss small cell treatment by cubicUpwindCPCFit, tweaks to figures

\documentclass{article}
\usepackage{showframe}
\usepackage[paperwidth=6.2in,paperheight=5.6in,margin=0in]{geometry}
\usepackage{mathptmx}%{times}
\usepackage{graphicx}
\usepackage{siunitx}
\usepackage[absolute,overlay]{textpos}
\usepackage{color}

\setlength{\parindent}{0pt}
\begin{document}
\TPMargin{3pt}
\textblockcolour{white}
\begin{textblock}{0.75}(1.4,0.5)
	\normalsize
	\centering
	(a)
\end{textblock}
\begin{textblock}{0.75}(1.2,6.5)
	\normalsize
	\centering
	(b)
\end{textblock}
\begin{textblock}{0.75}(9.05,6.5)
	\normalsize
	\centering
	(c)
\end{textblock}
\begin{textblock}{0.75}(1.2,11.3)
	\normalsize
	\centering
	(d)
\end{textblock}
\begin{textblock}{0.75}(9.05,11.3)
	\normalsize
	\centering
	(e)
\end{textblock}
\centering
\includegraphics[width=6in]{../gravityWaves-btf-250m-300dz/18000/thetaDiffOverview.pdf}
\vspace*{0.1in}
\includegraphics[width=3in]{../gravityWaves-btf-250m-300dz/18000/thetaDiffMountain.pdf}
\includegraphics[width=3in]{../gravityWaves-cutCell-250m-300dz/18000/thetaDiffMountain.pdf}
\includegraphics[width=3in]{../thermalAdvection-btf-250m-300dz/18000/tracerDiffMountain.pdf}
\includegraphics[width=3in]{../thermalAdvection-cutCell-250m-300dz/18000/tracerDiffMountain.pdf}

\newpage

250m GW

\includegraphics[width=3in]{../gravityWaves-btf-250m-250dz/18000/thetaDiffMountain.pdf}
\includegraphics[width=3in]{../gravityWaves-cutCell-250m-250dz/18000/thetaDiffMountain.pdf}

250m adv

\includegraphics[width=3in]{../thermalAdvection-btf-250m-250dz/18000/tracerDiffMountain.pdf}
\includegraphics[width=3in]{../thermalAdvection-cutCell-250m-250dz/18000/tracerDiffMountain.pdf}

\newpage
200m GW

\includegraphics[width=3in]{../gravityWaves-btf-250m-200dz/18000/thetaDiffMountain.pdf}
\includegraphics[width=3in]{../gravityWaves-cutCell-250m-200dz/18000/thetaDiffMountain.pdf}

200m adv

\includegraphics[width=3in]{../thermalAdvection-btf-250m-200dz/18000/tracerDiffMountain.pdf}
\includegraphics[width=3in]{../thermalAdvection-cutCell-250m-200dz/18000/tracerDiffMountain.pdf}

\newpage

150m GW

\includegraphics[width=3in]{../gravityWaves-btf-250m-150dz/18000/thetaDiffMountain.pdf}
\includegraphics[width=3in]{../gravityWaves-cutCell-250m-150dz/18000/thetaDiffMountain.pdf}

150m adv

\includegraphics[width=3in]{../thermalAdvection-btf-250m-150dz/18000/tracerDiffMountain.pdf}
\includegraphics[width=3in]{../thermalAdvection-cutCell-250m-150dz/18000/tracerDiffMountain.pdf}
\end{document}

>>>>

\documentclass{article}
\usepackage[paperwidth=6.2in,paperheight=2.6in,margin=0in]{geometry}
\usepackage{mathptmx}%{times}
\usepackage{graphicx}
\usepackage{siunitx}
\usepackage[absolute,overlay]{textpos}
\usepackage{color}
\usepackage{tikz}

\setlength{\parindent}{0pt}
\begin{document}
\TPMargin{3pt}
\begin{textblock}{0.75}(11.7,0.3)
\begin{tikzpicture}[
  scale=0.25
]
\draw [dashed, ultra thick] (0,0) -- (0,18);
\end{tikzpicture}
\end{textblock}
\centering
\includegraphics[width=6in]{../gravityWaves-btf-250m-50dz/18000/thetaDiffOverview.pdf} \\
\includegraphics[height=5in,angle=270]{thetaDiffOverview_theta_diff.eps}
\end{document}

### 2016-01-12 src/tf-cutcell-comparison/tf-cutcell-comparison.tex

The test was integrated forward by 5 hours using a timestep $\Delta t = \SI{100}{\second}$ on the BTF, SLEVE and cut cell grids, and a regular grid with flat terrain.
Maximum vertical velocities are presented in figure~\ref{fig:resting} and are similar on the BTF, SLEVE and cut cell grids.  The the largest vertical velocity of \SI{0.37}{\meter\per\second} was found on the BTF grid, compared with a maximum of \(\sim \SI{7}{\meter\per\second}\) found by \citet{klemp2011} using their improved horizontal pressure gradient formulation.
Unlike the result from \citet{klemp2011}, the SLEVE grid does not significantly reduce vertical velocities compared to the BTF grid.

>>>>

The test was integrated forward by 5 hours using a timestep $\Delta t = \SI{100}{\second}$ on the BTF, SLEVE and cut cell grids, and a regular grid with flat terrain.
Maximum vertical velocities are presented in figure~\ref{fig:resting} and are similar on the BTF, SLEVE and cut cell grids.  The largest vertical velocity of \SI{0.37}{\meter\per\second} was found on the BTF grid, compared with a maximum of \(\sim \SI{7}{\meter\per\second}\) found by \citet{klemp2011} using their improved horizontal pressure gradient formulation.
Unlike the result from \citet{klemp2011}, the SLEVE grid does not significantly reduce vertical velocities compared to the BTF grid.
